Data Specialization for a Staged Lambda Calculus

### 2015-02-16 paper/relatedwork.tex
first pass on example.tex
minor
finished a pass through 3.0-3.1

\section{Related Work}
There are several distinctive aspect of our work which we wish to compare and contrast.
Firstly, \lang\ is a simply typed lambda calculus 
with explicit stage annotations and a non-duplicating semantics.
Its statics and dynamics enforce no backward information flow,
which a is necessary property to define a splitting transformation
that is equivalent to the dynamics.
Finally, the splitting transformation itself operates on any well-typed open term in \lang.

@@

\subsection{Umut's notes}

---- DO NOT DELETE ----

Answer the following question for all of the related work:

\paragraph{Partial Evaluation.}
Online and offline.  Both approaches have their merits but online
partial evaluation can be too slow and if stopped lead to an
inefficient residual.  Offline partial evaluation is faster.  Also,
full self-application has only been achieved by offline partial
evaluation.

>>>>

\section{Related Work}

\crem{\emph{This text came from section 3:}
\lang\ is adapted from \cite{davies96}, and is very similar to other staged
languages considered in the partial evaluation and metaprogramming literature.
Unlike \cite{davies96}, we restrict \lang\ to exactly two stages, to simplify
our presentation of our stage-splitting algorithm. We expect that the techniques
presented here extend to more stages.
Our dynamic semantics for \lang\ differ from those given in \cite{davies96} and
the metaprogramming literature, because we avoid duplicating stage \bbtwo\
computations. In addition to being more efficient, our non-duplicating semantics
are easier to compare with our splitting algorithm, whose development and
exposition is our main goal.
}

There are several distinctive aspect of our work which we wish to compare and contrast.
Firstly, \lang\ is a simply typed lambda calculus 
with explicit stage annotations and a non-duplicating semantics.
Its statics and dynamics enforce no backward information flow,
which a is necessary property to define a splitting transformation
that is equivalent to the dynamics.
Finally, the splitting transformation itself operates on any well-typed open term in \lang.

@@

Online and offline.  Both approaches have their merits but online
partial evaluation can be too slow and if stopped lead to an
inefficient residual.  Offline partial evaluation is faster.  Also,
full self-application has only been achieved by offline partial
evaluation.

### 2015-02-16 paper/semantics.tex
note
finished a pass through 3.0-3.1

We express staged programs as terms in \lang, a typed, modal lambda calculus
with products, sums, and isorecursive types. This language is adapted from
\cite{davies96}, and is very similar to other staged languages considered in the
partial evaluation and metaprogramming literature, though it differs in three
key ways:

\begin{enumerate}

\item We restrict \lang\ to exactly two stages, to simplify our
presentation of our stage-splitting algorithm. We expect that the techniques
presented here extend to more stages.

\item Although \lang\ describes two-stage computations, its modal type system
has three distinct \emph{worlds} \bbonem, \bbonep, and \bbtwo. Essentially, this
further distinguishes those first-stage computations which are \emph{purely
first-stage} (\bbonep) or contain a \emph{mix} of code from each stage
(\bbonem). This distinction proves useful when defining our splitting algorithm.
%Only mixed code may transition to stage/world \bbtwo.

\item Our dynamic semantics for \lang\ differ from those given in \cite{davies96} and
the metaprogramming literature, because we avoid duplicating stage \bbtwo\
computations. In addition to being more efficient, our non-duplicating semantics
are easier to compare with our splitting algorithm, whose development and
exposition is our main goal.

\end{enumerate}

\crem{Do we actually want to make this ``more efficient'' claim?}

@@

All three worlds (\bbonem, \bbonep, and \bbtwo) contain product, sum, function,
and recursive types defined in the usual fashion. For example, given
two terms $\typesone{e_1}A$ and $\typesone{e_2}B$ at world \bbonem, one can form
the pair $\typesone{\texttt{<}e_1,e_2\texttt{>}}{A\times B}$ also at world
\bbonem.

Differing worlds (and hence, differing stages of computation) interact by means
of the $\fut$ and $\curr$ type formers. $\fut A$ is a type in world \bbonem\
which classifies second-stage computations of type $A$. Given a world \bbtwo\
term $e$ of type $A$, $\next{e}$ has type $\fut A$ at world
\bbonem. This essentially encapsulates $e$ as a computation to be evaluated in
the future---first-stage computations can shuffle it around, but not use its
result. The only way to use a $\fut A$ is to wrap it with $\prev$, yielding an
$A$ at world \bbtwo.

@@

@@

\crem{Need to explain why we chose letp and caseP as the abstractions.}

\crem{Need to explain why \bbonem\ doesn't have ints.}

>>>>

We express two-stage programs as terms in \lang, a typed, modal lambda calculus
with products, sums, and isorecursive types. Its type system has three distinct
\emph{worlds} \bbonep, \bbonem, and \bbtwo, which respectively classify
\emph{purely first-stage} computations (\bbonep); 
first-stage computations containing a \emph{mix} of code from each stage
(\bbonem); 
and second-stage computations (\bbtwo).

@@

All three worlds contain unit, product, sum, function, and recursive types
defined in the usual fashion. For example, given two terms $\typesone{e_1}A$ and
$\typesone{e_2}B$ at world \bbonem, one can form the pair
$\typesone{\texttt{<}e_1,e_2\texttt{>}}{A\times B}$ also at world \bbonem.

Differing worlds (and hence, differing stages of computation) interact by means
of the $\fut$ and $\curr$ type formers. $\fut A$ is a type in world \bbonem\
which classifies second-stage computations of type $A$. Given a term $e$ of type
$A$ at world \bbtwo, $\next{e}$ has type $\fut A$ at \bbonem. This essentially
encapsulates $e$ as a computation to be evaluated in the future---first-stage
computations can shuffle it around, but not use its result.  The only way to use
a $\fut A$ is to wrap it with $\prev$, yielding an $A$ at \bbtwo.

@@

Although our products and functions are restricted to types at the same world,
$\fut$ essentially allows us to build ``mixed-stage'' products and functions.
For example, quickselect is a function at world \bbonem\ which takes a
$\curr{\rm list} \times \fut\rmint$ (a purely-first-stage list and a
second-stage computation of an integer) to a $\fut\rmint$ (a second-stage
computation of an integer).

@@

An $e$ of type $\curr A$ at \bbonem\ can be unwrapped as an $A$ at \bbonep\
using the $\letp xe{e'}$ construct, which binds $\colpure xA$ in a \bbonem\ term
$e'$. This allows us to compute under $\curr$---for example, given a $\colmix
p{\curr(A\times B)}$, the term $\letp{x}{p}{\pure{\pio x}}$ computes its first
projection, of type $\curr A$.

Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we need the
$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
open on $\curr{A}$ and $\curr{B}$ respectively.

The example code in this paper makes liberal use of $\rmint$s and various
functions on these, as well as a function $\pause$ which takes a $\curr\rmint$
to a $\fut\rmint$.
